---
title: "ToothGrowth dataset analysis"
author: "Federico Calore"
date: "24 Oct 2015"
output: pdf_document
---
This document performs some exploratory data analysis on the ToothGrowth dataset provided with the base R package, and checks some relevant hypotheses based on the data.

From the [description](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html) in the dataset package:  

> **The Effect of Vitamin C on Tooth Growth in Guinea Pigs**  
> The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).

Dataset columns:

> variable | type    | description
> -------- | ------- | -----------
> $ len    | numeric | Tooth length
> $ supp   | factor  | Supplement type (VC or OJ).
> $ dose   | numeric | Dose in milligrams/day

## Initial data exploration

Load the ToothGrowth data from R datasets and perform a quick summary:
```{r}
data("ToothGrowth")
str(ToothGrowth)
summary(ToothGrowth)
```
-------

\pagebreak

The dataset contains results of 10 samples for each combination of *supplement type* and *dose*.
```{r, echo=FALSE}
cat("Count of observations:")
table(ToothGrowth$supp, ToothGrowth$dose)
```

We can quickly check the mean of the tooth growth subsetted by *supplement type* and *dose*:
```{r, comment=""}
# mean of length by supplement type
tapply(ToothGrowth$len, INDEX = ToothGrowth$supp, FUN = mean)

# mean of length by dose
tapply(ToothGrowth$len, INDEX = ToothGrowth$dose, FUN = mean)

# mean of length by combinations of supplement and dose
xtabs(len/10 ~ supp + dose, data = ToothGrowth)
```

We can also explore the results with some boxplots to highlight differences in distribution of the tooth length variable by the *supp* and *dose* factors:  

```{r, echo=FALSE}
par(mfrow = c(1,2))
boxplot(len ~ supp, data = ToothGrowth, ylab = "length by supplement type")
boxplot(len ~ dose, data = ToothGrowth, ylab = "length by dose")
```

-------

\pagebreak

## Hypothesis testing

### Hypothesis testing on supplement type
Looking at the boxplot on the left, it looks like the ascorbic acid is less effective than the orange juice.
We want to check if we have enough statistical evidence to support this. We can formally define the alternatives as follows:

> $H_0:   \mu_{OJ} = \mu_{VC}$  
> $H_1:   \mu_{OJ} \neq \mu_{VC}$  

We will use a two sided t-test to verify the interval of confidence in order to reject the *null* hypothesis H~0~.  
We use unpaired observations, since the guinea pigs in one group didn't have anything to do with the ones in the other group, and assume unequal variances.

```{r, collapse=TRUE}
t <- t.test(len ~ supp, data = ToothGrowth)
t$conf[1:2]
t$p.value
```

The *p-value* is greater than 0.05 and the confidence interval contains zero, therefore we don't have enough statistical evidence to reject the *null* hypothesis with 95% confidence.

If we chose a relaxed *90% interval* for the test, or we used a *one-sided* hypothesis (*greater than*), the interval would not include zero, and we would reject the *null* hypothesis in favor of the alternative:

```{r}
 # 90% confidence interval
t.test(len ~ supp, data = ToothGrowth, conf.level = 0.90)$conf[1:2]

# one sided alternative hypothesis
t.test(len ~ supp, data = ToothGrowth, alternative = "greater")$conf[1:2]
```

### Hypothesis testing on doses
We have three levels for the doses, so we will have to compare them in pairs. The *null* hypothesis will assume an equal mean across levels, and the alternative will assume there is a difference.

As in the previous example, we will use two sided t-tests with unpaired observations and unequal variances.
The result is formatted in a matrix to improve readability.

```{r, echo=FALSE}
t1 <- with(ToothGrowth, t.test(len[dose == 0.5], len[dose == 1])) # .5 vs 1
t2 <- with(ToothGrowth, t.test(len[dose == 0.5], len[dose == 2])) # .5 vs 2
t3 <- with(ToothGrowth, t.test(len[dose == 1], len[dose == 2]))   #  1 vs 2

data.frame(
  comparisons = c(".5 vs 1", ".5 vs 2", " 1 vs 2"),
  interval = rbind(
    round(t1$conf[1:2], digits = 3),
    round(t2$conf[1:2], digits = 3),
    round(t3$conf[1:2], digits = 3)
  ),
  p.values = rbind(
    round(t1$p.value, digits = 5),
    round(t2$p.value, digits = 5),
    round(t3$p.value, digits = 5)
  )
)
```

All three intervals are entirely below zero, so we would reject the *null* hypothesis in favor of the alternative in all cases. That implies *0.5 < 1 < 2*, or in other words, the more vitamin C, the stronger the effect.
